Algebraic Groups\vspace{-1ex}

### 2020-02-03 Algebraic Geometry/Algebraic-Geometry.tex
update

>>>>

\section{January 31, 2020}
Today we are going to talk about fiber products. Max thinks they should be ``fibered products'' but he's wrong.

\subsection{Fiber products are real good}
\begin{defn}
	A fiber product $X\times_Z Y$ is the limit of $X\to Z\leftarrow Y$.
	You know the drill.
\end{defn}
Equivalently the fiber product of schemes is the fiber product of the functor $\Sch\op\to \Set$ that it represents.
The reason this is worth mentioning is that one can use the fuctor of points perspective to define this as a fiber product 
on sets!

\begin{thm}
	Fiber products exist in $\Sch$.
\end{thm}
\begin{prf}
	We gonna do this.
	\begin{defn}
		A functor $\calF:\Sch\op\to\Set$ is a \textbf{sheaf} if for all $W\in\Sch$, the induced functor 
		$\calF_W:\mathbf{Open}(W)\op\to\Set$ is a sheaf.
	\end{defn}
	Kinda obvious. Then the idea here is that the Yoneda embedding actually maps into sheaves on schemes (not just presheaves). Furthermore 
	since the category of sheaves has all limits (CHECK THIS! It's easy), this means we are just trying to show that the sheaf $h_X\times_{h_Y}h_Z$ is representable 
	by a scheme.

	\begin{defn}
		A natural transformation $\calG\to \calF$ in $\Func(\Sch\op,\Set)$ is an \textbf{open subfunctor} if, 
		for all schemes $W$ and all $h_W\to F$, the pullback $G\times_F h_W\to h_W$ is representable by 
		an open immersion $h_U\to h_W$ ($U\subseteq W$ open).
	\end{defn}
	We're trying to define a thing like a topology on the category of schemes.
	\begin{defn}
		Given a fucntor $\calF\in\PreSh(\Sch)$, an \textbf{open covering} is a collection of open 
		subfunctors $\calG_i\hookrightarrow\calF$ such that for all $W\in\Sch$ $h_{U_i}=\calG_i\times\calF h_W\to h_W$.
	\end{defn}
	\begin{prop}
		A sheaf $\calF:\PreSh(\Sch)$ is representable by a scheme if and only if $\calF$ admits an open covering by functors 
		representable by affine schemes.
	\end{prop}
	You can try to prove it! It also is done in Demazure and Gabriel although I am pretty sure that is in French and/or inscrutable. 
	Max says it's not too hard, though. You can kinda follow the argument in Hartshorne for the analogous thought coming from the category of locally ringed spaces.

	\begin{lem}
		Let $U,V,W$ be opens in $X,Y,Z$, resp. such that $f(U)\subset W\supset g(V)$. Then $h_U\times_{h_W}h_V\to h_X\times_{h_Z}h_Y$ is an open subfunctor.
	\end{lem}
	Again easy to check. This gives us (from affine coverings of $X,Y,Z$) an open covering by $h_{\Spec A}\times_{h_{\Spec C}}h_{\Spec B}$.

	From here we can use (among other things) that the restricted Yoneda $\Sch\to\PreSh(\Aff)$ is fully faithful.
\end{prf}
You can read all of today's lecture in Vakil chapter 9.
\section{February 3, 2020}
Today Adam is giving the lecture! The title is

\subsection{Category theory for algebraic geometry}
This lecture is assuming that we know about categories, functors, and natural transformations.
We will be focusing on limits and colimits.

A motivating question: say we have an open cover $\cup U_i=X$. How can we relate the sheaves on $X$ to those on the $U_i$? Saw we have sheaves $\calF_i$ on $U_i$ along with 
isomorphisms $\phi_{ij}:\calF_i|_{U_i\cap U_j}\to \calF_j|_{U_i\cap U_j}$ (along with a cocycle condition we'll discuss later). How can a sheave $\calF$ on $X$ 
be related to these?

Take an open $V\subseteq X$. Then for all $i,j,$ we have maps
\begin{figure}[h]
	\centering
	\begin{tikzcd}
		& \calF(V)\ar[dl]\ar[dr] &\\
		\calF(V\cap U_i)\ar[d] & & \calF(V\cap U_j)\ar[d]\\
		\calF_i(V\cap U_i)\ar[d] & & \calF_j(V\cap U_j)\ar[d]\\
		\calF_i(V\cap U_i\cap U_j)\ar[rr,"\varphi_{ij}"] & & \calF_j(V\cap U_i\cap U_j)
	\end{tikzcd}
\end{figure}

The gluing idea is to define $\calF(V)$ to be the limit of the bottom portion of the diagram. To do this in general,
let $\calC$ be a category and $K:\calJ\to\calC$ for some other category $\calJ$. Define a cone over $K$ with summit $c\in \calC$ 
to be the usual thing. Same for a cone under a functor. Then limits and colimits. Most of this is 
basically the same as Emily Riehl.

He uses the fact that $\operatorname{cone}(-,K):\calC\op\to\Set$ (this assumes $\calJ$ is small) is a functor,
and furthermore that 
\[\Hom_\calC(-,\lim K)\simeq \operatorname{cone}(-,K)\]
which is essentially a restatement of the universal property of a limit.

Some examples:
\begin{itemize}
	\item In $\Set$, the limit of $A\hookrightarrow X\hookleftarrow B$ is $A\cap B$.
	\item Equalizers are limits!
\end{itemize}

### 2020-02-03 Algebraic Groups/Algebraic-Groups.tex
update

>>>>

\section{January 31, 2020}
Recall that we had $\Lie G$ as well as bijections between a couple different things.
Today we are going to talk more about that. We scribbled out some easy computations for why (b) and (c) are equivalent.
The idea here is that any map $A\to k\oplus k\tau$ can be written as $(\varphi,d)$ where $d$ is a $\varphi$ derivation.

To show the equivalence of $(a)$ and $(b)$, you use the maps $D\mapsto \varepsilon D$ and the inverse $d\mapsto (\id\otimes d)\circ\Delta$. One needs to check that you get derivations and that the other thing 
is left-invariant. To show that they are inverses, consider 
\[\varepsilon(\id\bar\otimes d)\circ \Delta(x)=\sum\varepsilon(a')d(a'')=d(\sum \varepsilon(a')a'')=d(a).\]
Leave the other direction as an exercise.

Let's check left invariance: we need to check (by the discussion the other day) that $\Delta\circ D=(d\otimes D)\circ\Delta.$ Checking this si doing a little diagram chase.
Part of the diagram is associativity.

\subsection{Lie structures}
A natural question to ask: if we have this isomorphism in this last lemma, what are the brackets on the other two objects?

For the first, let $d_1,d_2\in\Der_\varepsilon(A,k)$. Let $D_i=(\id\bar\otimes d_i)\circ\Delta$. Do some work and chase the diagrams for $[D_1,D_2]$!
Then you find that 
\[[d_1,d_2]=(d_1\bar\otimes d_2-d_2\bar\otimes d_1)\circ\Delta\]
which really underlines that we need the group/Hopf structure on these things to really be able to do this! This is what makes these equivalences work!

For the kernel $\ker p_\ast$, take two elements $g_1=(\varepsilon, d_1)\in k[\tau_1]$ and $g_2=(\varepsilon, d_2)\in k[\tau_2]$ where we 
are using different nilpotents so that we have enough degrees of freedom. Consider $R=k[\tau_1,\tau_2]$ and consider the inclusion of $g_1$ and $g_2$ in $G(R)=\Hom(A,R)$.

So we think of $g_1=\varepsilon+d_1\tau_1$ and $g_2=\varepsilon+d_2\tau_2$ and we compute the multiplication as:
\[g_1g_2=\varepsilon+d_1\tau_1+d_2\tau_2+((d_1\bar\otimes d_2)\circ\Delta)\tau_1\tau_2\]
and coompute $g_2g_1$ as well. Then we really don't have additive inverses in $G$, but we can compute 
\[g_1g_2=(\varepsilon+[d_1,d_2]\tau_1\tau_2)g_2g_1.\]

How do we use this in practice? Consider $\GL_n(S)$. The Lie algebra consists of the matrices $M\in\GL_n(k[\tau])$.
Each such matrix can be written as $M_1+M_2\tau$. But we want the condition that the projection onto $\GL_n(k)$ is the identity,
so really we have 
\[\Lie\GL_n = \{I+M\tau\}\subset\GL_n(k[\tau])\]
but any such matrix is invertible if $M$ is, so it is no further condition. This gets you your isomorphism with all of $\gl_n$.

\section{February 3rd, 2020}
Working from last time, we have some corollaries:
\begin{cor}
	Let $f:H\to G$ be a map. Then 
	\begin{itemize}
		\item $f$ induces a map $\Lie H\to \Lie G$ ($\Lie$ is functorial).
		\item If $f$ is inclusion, then $\Lie H\hookrightarrow\Lie G$
	\end{itemize}
\end{cor}
\begin{rmk}
	This gives us the full equivalences 
	\[\Lie G\simeq\Der_\varepsilon(k[G],k)\simeq (I/I^2)^\ast\simeq T_e G.\]
\end{rmk}
and from this we get 
\begin{cor}
	$G$ is smooth (or reduced) if and only if $\dim G=\dim_k \Lie G$.
\end{cor}
where the above is true since $\dim_k\Lie G=\rank \Omega_{k[G]}$.

\subsection{Examples}
We talked about $\GL_n$ last time. Also if we have $I+\tau_1M_1,I+\tau_2 M_2\in\Lie \GL_n=\gl_n$, the bracked can quickly be seen to be 
\[I+\tau_1\tau_2[M_1,M_2]\]
as one would hope.

Now consider $\SL_n$. Then $\Lie\SL_n\subset\gl_n$. To analyze this, consider $\SL_n(k[\tau])$ 
and we are looking for the ones with determinant one. But 
\[\det(I+\tau M)=p(\tau)\]
where $p$ is the characteristic polynomial for $I+\tau M$ over $k$. But the terms above the linear term die and the linear term is the trace (of $M$):
\[\det(I+\tau M)=1+(\tr M)\tau=1\]
so the condition is $\tr M=0$.

\subsection{Positive characteristic}
Recall that $\Ga$ is the group scheme represented by $k[t]$ where $t$ is primitive. $\Gm$ is 
represented by $k[t,t^{-1}]$ where $t$ is grouplike.

These are both one-dimensional smooth group schmemes, so the Lie algebras $\frakg_a=\Lie \Ga$ and $\frakg_m=\Lie\Gm$
are both one-dimensional. If $\ch k=0$, then, there is only one Lie structure! But what if not?
Assume for now $\ch k\ne 2$.

Do the following problem! It's not hard.
\begin{prob}
	Let $\ch k=p$. Take a derivation $D\in \Der_k(A,A)$ and show that the $p^{th}$ power of $D$ is again a derivation.
\end{prob}

This give rise to a ``$p^{th}$ power map'' which gives rise to a \textbf{restricted Lie algebra}. See Weibel for a good 
introduction (he had to really learn it well to write about it so it is very clear).

Go back to the case of $\Ga$ and $\Gm$. $\frakg_a=\Der_\varepsilon(k[t],k)=\langle d_t=\frac{d}{dt}\rangle$
such that $d_tt=1$.

Then $D_t=(\id\bar\otimes d_t)\circ\Delta$ as the corresponding derivation on $k[G]$, and we can compute 
\begin{align*}
	D_t(t)&= (\id\bar\otimes d_t)\circ \Delta(t)\\
	&= (\id\bar\otimes d_t)(1\otimes t+t\otimes 1)=1
\end{align*}
and so $D_t^2(t)=0$. Thus $D_t^p(t)=0$, so $\frakg_a$ is a trivial $p$ Lie algebra: $\frakg_a=\langle d_t\rangle$
where $d_t^p=0$.

In the case of $\Gm$, we can compute 
\[D_t(t)=t\]
so then $D_t^p=D_t$, so we get $\frakg_m=\langle d_t\rangle$ with $d_t^p=d_t$. The punchline here is 
that the two infinitessimal theories, at first glance, appear to coincide although they actually vary 
in the right light. $\frakg_a\cong\frakg_m$ as Lie algebras, but not as \textit{restricted} Lie algebras.

\subsection{Frobenius Kernels}
For reference, this can be found in Jantzen part I section 7. Some of the proof will be omitted, but can be found there.

Let $\ch k=p>0$ and let $G$ be an algebraic group scheme over $k$. We assume here that $k=\bar k$, but it suffices that 
$k$ is perfect. This fixes some problems that arise. Recall the Frobenius map $F:\A n\to \A n$
that sends $(a_1,\dots,a_n)\mapsto(a_1^p,\dots,a_n^p)$. The \textbf{Frobenius twist} of $G$ is the image 
$G^{(1)}$ of $G$ under the Frobenius twist.

Then we can define the \textbf{Frobenius kernel}
\[G_{(1)}\ker F\hookrightarrow G\xrightarrow{F} G^{(1)}\]
which one can show is a connected ($k[G_{(1)}]$ is local) and finite ($k[G_{(1)}]$ is finite dimensional) and a group scheme.

Is the Frobenius kernel smooth? Usually no! The finite dimensionality of $k[G_{(1)}]$ means the dimension of $G$ is zero,
so $\Lie G$ is trivial. But this happens only if $G$ is trivial.

An example: $\Ga$. Consider the Frobenius map $F:\Ga\to \Ga$ via $r\mapsto r^p$. Then the kernel 
is the collection of $p$-nilpotents, 
\[{\Ga}_{(1)}(R)=\{r\in R|r^p=0\}\]
and one can compute that 
\[k[{\Ga}_{(1)}]\cong k[t]/t^p\]
with $t$ being a primitive element. We call ${\Ga}_{(1)}=\alpha_p$.

When we switch the picture to $\Gm$, we are now looking at $p^{th}$ roots of unity! The algebra is 
$k[t,t^{-1}]/(t^p-1)=k[t]/(t^p-1)\simeq k[t-1]/(t-1)^p\simeq k[u]/u^p$. So the coordinate algebra is isomorphic 
to that in the last case, but with a different coproduct.
